---
title: "Plank study sessions: experimental wide data pipeline"
subtitle: "From raw export to `w241_experiment_wide.csv` and codebook"
author: "W241 Plank Study"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  collapse = TRUE
)
```

## Purpose and outputs

This document **documents and runs** the transformation from the app export **`W241 Plank Study Data - sessions.csv`** (long format: one row per participant-session) into:

1. **`w241_experiment_wide.csv`** — one row per `participant_id`, analysis-ready survey fields plus participant demographics/metadata, with no email or name.
2. **`w241_experiment_codebook.csv`** — maps integer codes to labels for columns that are coded (survey and process items).

Design choices (high level):

- **Music labels:** raw `trance` / `pop` are renamed to **`high_tempo`** and **`low_tempo`** for clarity in analysis files.
- **Completion:** **`completed_both`** is `1` if both sessions exist after deduplication, **`0` if only one session**. Partial completers are **kept** in the wide file (not dropped).
- **Time:** Raw submission timestamps are used **only** to compute **`elapsed_sessions_hours`** between session 1 and session 2 (UTC). They are **not** written to the wide CSV (privacy and simplicity).
- **Participant metadata:** non-PII participant fields such as **`age`**, **`gender`**, **`locale`**, and planned-session metadata are merged from **`W241 Plank Study Data - participants.csv`** by `participant_id`.
- **Survey JSON:** Parsed to **integer** columns; yes/no \(\rightarrow\) 0/1; `volume_clear` adds **somewhat = 2**; free-text fields become **0/1 presence** columns (text itself is not exported).

Run the whole document with **Knit** (or `rmarkdown::render()`) to regenerate the CSVs. Input/output paths are built from **`knitr::current_input()`** so **`W241 Plank Study Data - sessions.csv`** is read from the same folder as this Rmd, even when the process cwd is elsewhere (no reliance on the Knit working directory).

---

## Libraries and file paths

```{r libraries}
library(data.table)
library(jsonlite)
# knitr loaded in setup

.rmd_dir <- {
  fn <- tryCatch(knitr::current_input(), error = function(e) "")
  if (is.null(fn) || !nzchar(fn)) {
    getwd()
  } else {
    dirname(normalizePath(fn, winslash = "/"))
  }
}
SRC <- file.path(.rmd_dir, "W241 Plank Study Data - sessions.csv")
PARTICIPANTS_SRC <- file.path(.rmd_dir, "W241 Plank Study Data - participants.csv")
OUT_WIDE <- file.path(.rmd_dir, "w241_experiment_wide.csv")
OUT_CODEBOOK <- file.path(.rmd_dir, "w241_experiment_codebook.csv")

AUDIO_MAP <- c(trance = "high_tempo", pop = "low_tempo")

YN_FIELDS <- c("headphones", "plank_pause")
VOLUME_FIELD <- "volume_clear"
TEXT_FIELDS <- c("comments", "plank_pause_detail")
NUMERIC_FIELDS <- c(
  "energy_pre_plank", "motivation_pre_plank", "music_liking",
  "rpe", "music_effect", "music_effect_overall",
  "instructions_ease", "overall_experience"
)
```

**`AUDIO_MAP`:** maps the values stored in the export to analysis labels. Any other `audio_track` value causes an explicit error so bad data are not silently relabeled.

**Field lists** drive how each JSON key is encoded (boolean, ordinal volume, numeric Likert, text presence, or string categorical).

---

## Helper functions: JSON parsing

Exports from spreadsheets often embed JSON with **doubled double-quotes** (`""` inside the field) instead of backslash-escaped quotes. `jsonlite::fromJSON` expects standard JSON, so we **replace `""` with `"`** before parsing. Malformed or empty strings return `NULL` and become missing values downstream.

```{r fn-parse}
parse_obj <- function(txt) {
  if (is.na(txt) || !nzchar(trimws(as.character(txt)))) return(NULL)
  txt <- gsub('""', '"', txt, fixed = TRUE)
  out <- tryCatch(
    fromJSON(txt, simplifyVector = TRUE),
    error = function(e) NULL
  )
  if (is.null(out) || length(out) == 0L) return(NULL)
  if (!is.list(out)) return(NULL)
  out
}
```

---

## Helper functions: encoding rules

### Yes / no and volume

`headphones` and `plank_pause` are coded **0 = no**, **1 = yes** (case-insensitive; common variants like `true` / `false` accepted). **`volume_clear`** is **0 = no**, **1 = yes**, **2 = somewhat** (a third level that appears in the forms).

```{r fn-yn-volume}
norm_yn <- function(x) {
  if (is.null(x) || (length(x) == 1L && is.na(x))) return(NA_integer_)
  v <- tolower(trimws(as.character(x[[1L]])))
  if (!nzchar(v)) return(NA_integer_)
  if (v %in% c("yes", "y", "true", "1")) return(1L)
  if (v %in% c("no", "n", "false", "0")) return(0L)
  NA_integer_
}

norm_volume <- function(x) {
  if (is.null(x) || (length(x) == 1L && is.na(x))) return(NA_integer_)
  v <- tolower(trimws(as.character(x[[1L]])))
  if (!nzchar(v)) return(NA_integer_)
  if (v == "yes") return(1L)
  if (v == "no") return(0L)
  if (v == "somewhat") return(2L)
  NA_integer_
}
```

### Numeric Likert fields

Listed keys are coerced to **integers** (rounded) for scales such as RPE and pre-task motivation.

```{r fn-numeric}
as_int_num <- function(x) {
  if (is.null(x) || (length(x) == 1L && is.na(x))) return(NA_integer_)
  z <- suppressWarnings(as.integer(round(as.numeric(x[[1L]]))))
  if (length(z) != 1L || is.na(z)) NA_integer_ else z
}

text_present <- function(x) {
  if (is.null(x)) return(NA_integer_)
  s <- trimws(as.character(x))
  if (length(s) != 1L) return(NA_integer_)
  if (!nzchar(s) || s == "NA") return(0L)
  1L
}
```

### Participant metadata

The sessions export provides the repeated-measures outcomes and survey JSON, while the participant export provides baseline non-PII fields keyed by `participant_id`. We merge those fields into the final wide file so later analysis does not have to re-open the participant CSV.

```{r fn-participants}
load_participants <- function(path) {
  if (!file.exists(path)) {
    stop("Required participants file not found: ", path)
  }
  p <- fread(path)
  stopifnot("id" %in% names(p))
  setnames(p, "id", "participant_id")

  keep <- intersect(
    c("participant_id", "locale", "age", "gender", "group_index",
      "group_label", "sessions_completed", "registered_at",
      "session1_planned_at", "session2_planned_at"),
    names(p)
  )
  p <- copy(p[, keep, with = FALSE])

  if ("age" %in% names(p)) {
    p[["age"]] <- suppressWarnings(as.numeric(p[["age"]]))
    setnames(p, "age", "participant_age")
  }
  if ("gender" %in% names(p)) {
    p[["gender"]] <- trimws(tolower(as.character(p[["gender"]])))
    p[!nzchar(p[["gender"]]), "gender"] <- NA_character_
    setnames(p, "gender", "participant_gender")
  }
  if ("locale" %in% names(p)) setnames(p, "locale", "participant_locale")
  if ("group_index" %in% names(p)) setnames(p, "group_index", "participant_group_index")
  if ("group_label" %in% names(p)) setnames(p, "group_label", "participant_group_label")
  if ("sessions_completed" %in% names(p)) {
    p[["sessions_completed"]] <- suppressWarnings(as.integer(p[["sessions_completed"]]))
    setnames(p, "sessions_completed", "participant_sessions_completed")
  }
  if ("registered_at" %in% names(p)) {
    p[["registered_at"]] <- as.POSIXct(p[["registered_at"]], tz = "UTC", format = "%Y-%m-%dT%H:%M:%OSZ")
    setnames(p, "registered_at", "participant_registered_at")
  }
  if ("session1_planned_at" %in% names(p)) {
    p[["session1_planned_at"]] <- as.POSIXct(p[["session1_planned_at"]], tz = "UTC", format = "%Y-%m-%dT%H:%M:%OSZ")
    setnames(p, "session1_planned_at", "participant_session1_planned_at")
  }
  if ("session2_planned_at" %in% names(p)) {
    p[["session2_planned_at"]] <- as.POSIXct(p[["session2_planned_at"]], tz = "UTC", format = "%Y-%m-%dT%H:%M:%OSZ")
    setnames(p, "session2_planned_at", "participant_session2_planned_at")
  }

  unique(p, by = "participant_id")
}
```

### Categorical strings and codebook accumulation

Any JSON key **not** in the boolean, volume, numeric, or text lists is treated as a **categorical string**. Distinct strings are collected into a global catalog per key (sorted alphabetically), assigned integer codes `1, 2, …`, and described in the **codebook** so labels are recoverable.

```{r fn-catalog-encode}
catalog <- new.env(parent = emptyenv())
codebook_rows <- list()

add_codebook <- function(variable, code, label) {
  codebook_rows[[length(codebook_rows) + 1L]] <<- data.table(
    column_name = variable,
    code_value = code,
    value_label = label
  )
}

encode_scalar <- function(key, val, colname) {
  if (key %in% TEXT_FIELDS) {
    p <- text_present(val)
    return(list(col = paste0(colname, "_present"), val = p))
  }
  if (key %in% YN_FIELDS) {
    return(list(col = colname, val = norm_yn(val)))
  }
  if (identical(key, VOLUME_FIELD)) {
    return(list(col = colname, val = norm_volume(val)))
  }
  if (key %in% NUMERIC_FIELDS) {
    return(list(col = colname, val = as_int_num(val)))
  }
  if (is.null(val) || (length(val) == 1L && is.na(val))) {
    return(list(col = colname, val = NA_integer_))
  }
  lab <- trimws(as.character(val[[1L]]))
  if (!nzchar(lab)) return(list(col = colname, val = NA_integer_))

  ck <- paste0("cat__", key)
  if (!exists(ck, envir = catalog, inherits = FALSE)) {
    assign(ck, character(), envir = catalog)
  }
  lev <- get(ck, envir = catalog)
  if (!lab %in% lev) {
    lev <- sort(unique(c(lev, lab)))
    assign(ck, lev, envir = catalog)
  }
  lev <- get(ck, envir = catalog)
  code <- match(lab, lev)
  list(col = colname, val = as.integer(code))
}
```

### Flattening pre/post JSON per session

Each session contributes **`pre_*_s1`** / **`post_*_s1`** (or **`_s2`**) columns. Text fields (`comments`, `plank_pause_detail`) become **`…_present_s1`** style columns with values 0/1 only.

```{r fn-flatten}
flatten_json_block <- function(obj, prefix, session_num) {
  if (is.null(obj)) return(list())
  sfx <- paste0("_s", session_num)
  out <- list()
  for (key in names(obj)) {
    val <- obj[[key]]
    base_col <- paste0(prefix, "_", key, sfx)
    if (key %in% TEXT_FIELDS) {
      enc <- encode_scalar(key, val, paste0(prefix, "_", key))
      nm <- paste0(enc$col, sfx)
      out[[nm]] <- enc$val
      next
    }
    enc <- encode_scalar(key, val, base_col)
    out[[enc$col]] <- enc$val
  }
  out
}
```

---

## First pass: discover all categorical levels

Before assigning final integer codes, we scan **every row** of the deduplicated long table and every parsed pre/post object so that **all** string levels for each categorical key are known. That keeps codes stable when we encode row by row (no surprise new level mid-file).

```{r fn-build-catalog}
build_catalogs_pass <- function(dt) {
  for (i in seq_len(nrow(dt))) {
    pre <- parse_obj(dt$pre_task_answers[i])
    post <- parse_obj(dt$post_task_answers[i])
    for (obj in list(pre, post)) {
      if (is.null(obj)) next
      for (key in names(obj)) {
        if (key %in% c(YN_FIELDS, VOLUME_FIELD, NUMERIC_FIELDS, TEXT_FIELDS)) next
        val <- obj[[key]]
        if (is.null(val)) next
        lab <- trimws(as.character(val[[1L]]))
        if (!nzchar(lab) || is.na(lab)) next
        ck <- paste0("cat__", key)
        if (!exists(ck, envir = catalog, inherits = FALSE)) assign(ck, character(), envir = catalog)
        cur <- get(ck, envir = catalog)
        if (!lab %in% cur) assign(ck, sort(unique(c(cur, lab))), envir = catalog)
      }
    }
  }
}

register_cat_codebooks_for_session <- function(keys, prefix, session_num) {
  sfx <- paste0("_s", session_num)
  for (key in keys) {
    if (key %in% c(YN_FIELDS, VOLUME_FIELD, NUMERIC_FIELDS, TEXT_FIELDS)) next
    ck <- paste0("cat__", key)
    if (!exists(ck, envir = catalog, inherits = FALSE)) next
    lev <- get(ck, envir = catalog)
    colnm <- paste0(prefix, "_", key, sfx)
    for (i in seq_along(lev)) {
      add_codebook(colnm, i, lev[i])
    }
  }
}
```

---

## Load long data and participant metadata, map audio, deduplicate

**Steps:**

1. Read CSV; require core column names.
2. Read the participant export keyed by **`participant_id`** and keep non-PII metadata for the final wide file.
3. Drop **`email`** and **`contact_sheet_url`** from the sessions table so they never appear in outputs.
4. Normalize and map **`audio_track`** via `AUDIO_MAP`; stop if unknown values appear.
5. Parse **`submitted_at`** as UTC (ISO format with `Z`); fallback parser if the first attempt yields all `NA`.
6. Sort by participant, session, time; for each (**`participant_id`**, **`session_num`**) keep the **last** row (latest submission) to resolve duplicate submissions.

```{r load-clean}
dt <- fread(SRC)
participants <- load_participants(PARTICIPANTS_SRC)
stopifnot(all(c(
  "participant_id", "session_num", "audio_track",
  "plank_duration_sec", "pre_task_answers", "post_task_answers", "submitted_at"
) %in% names(dt)))

if ("email" %in% names(dt)) dt[, email := NULL]
if ("contact_sheet_url" %in% names(dt)) dt[, contact_sheet_url := NULL]

dt[, audio_track := trimws(tolower(as.character(audio_track)))]
bad_audio <- unique(dt[!audio_track %in% names(AUDIO_MAP), audio_track])
if (length(bad_audio)) {
  stop("Unknown audio_track value(s): ", paste(bad_audio, collapse = ", "))
}
dt[, audio_track := unname(AUDIO_MAP[audio_track])]

dt[, submitted_at := as.POSIXct(submitted_at, tz = "UTC", format = "%Y-%m-%dT%H:%M:%OSZ")]
if (all(is.na(dt$submitted_at))) {
  dt[, submitted_at := as.POSIXct(submitted_at, tz = "UTC")]
}

setorder(dt, participant_id, session_num, submitted_at)
dt <- dt[, .SD[.N], by = .(participant_id, session_num)]

build_catalogs_pass(dt)
```

After this chunk, **`dt`** is the canonical long table used for the rest of the pipeline.

---

## Build session-specific tables and merge to wide

For each session number (1 and 2), we:

1. Collect which JSON keys appear in pre/post for that session slice (varies: session 1 pre often has demographics; session 2 pre is shorter).
2. Register **codebook** rows for categoricals, yes/no, volume, and text-presence columns (some codebook rows may reference columns that do not appear in the final wide file if a key never occurs for that session; those are filtered out at export).
3. For each participant row in that session, flatten pre/post JSON into a single wide row fragment and **`rbindlist`** with `fill = TRUE`.

We then **outer-join** on **`participant_id`** so every participant appears once: those with only session 1 or only session 2 get **`NA`** for the other session’s columns.

```{r process-session}
process_session <- function(dt, sn) {
  d <- dt[dt$session_num == sn]
  if (nrow(d) == 0L) {
    return(data.table(participant_id = character()))
  }

  keys_pre <- unique(unlist(lapply(seq_len(nrow(d)), function(i) {
    o <- parse_obj(d$pre_task_answers[i])
    if (is.null(o)) character() else names(o)
  })))
  keys_post <- unique(unlist(lapply(seq_len(nrow(d)), function(i) {
    o <- parse_obj(d$post_task_answers[i])
    if (is.null(o)) character() else names(o)
  })))
  register_cat_codebooks_for_session(keys_pre, "pre", sn)
  register_cat_codebooks_for_session(keys_post, "post", sn)

  for (fld in YN_FIELDS) {
    add_codebook(paste0("post_", fld, "_s", sn), 0L, "no")
    add_codebook(paste0("post_", fld, "_s", sn), 1L, "yes")
    add_codebook(paste0("pre_", fld, "_s", sn), 0L, "no")
    add_codebook(paste0("pre_", fld, "_s", sn), 1L, "yes")
  }
  add_codebook(paste0("post_", VOLUME_FIELD, "_s", sn), 0L, "no")
  add_codebook(paste0("post_", VOLUME_FIELD, "_s", sn), 1L, "yes")
  add_codebook(paste0("post_", VOLUME_FIELD, "_s", sn), 2L, "somewhat")
  add_codebook(paste0("pre_", VOLUME_FIELD, "_s", sn), 0L, "no")
  add_codebook(paste0("pre_", VOLUME_FIELD, "_s", sn), 1L, "yes")
  add_codebook(paste0("pre_", VOLUME_FIELD, "_s", sn), 2L, "somewhat")

  for (fld in TEXT_FIELDS) {
    add_codebook(paste0("pre_", fld, "_present_s", sn), 0L, "empty or missing")
    add_codebook(paste0("pre_", fld, "_present_s", sn), 1L, "non-empty text (not exported)")
    add_codebook(paste0("post_", fld, "_present_s", sn), 0L, "empty or missing")
    add_codebook(paste0("post_", fld, "_present_s", sn), 1L, "non-empty text (not exported)")
  }

  rows <- vector("list", nrow(d))
  for (i in seq_len(nrow(d))) {
    pre <- parse_obj(d$pre_task_answers[i])
    post <- parse_obj(d$post_task_answers[i])
    L <- c(
      flatten_json_block(pre, "pre", sn),
      flatten_json_block(post, "post", sn)
    )
    rows[[i]] <- c(
      list(
        participant_id = d$participant_id[i],
        plank_duration_sec = d$plank_duration_sec[i],
        audio_track = d$audio_track[i],
        submitted_at = d$submitted_at[i]
      ),
      L
    )
  }
  rbindlist(lapply(rows, function(r) as.data.table(r)), fill = TRUE)
}

s1 <- process_session(dt, 1L)
s2 <- process_session(dt, 2L)

setnames(s1, c("plank_duration_sec", "audio_track", "submitted_at"),
         c("plank_s1", "audio_s1", "submitted_at_s1"), skip_absent = TRUE)
setnames(s2, c("plank_duration_sec", "audio_track", "submitted_at"),
         c("plank_s2", "audio_s2", "submitted_at_s2"), skip_absent = TRUE)

all_pid <- unique(c(s1$participant_id, s2$participant_id))
wide <- data.table(participant_id = all_pid)
wide <- merge(wide, s1, by = "participant_id", all.x = TRUE)
wide <- merge(wide, s2, by = "participant_id", all.x = TRUE)
wide <- merge(wide, participants, by = "participant_id", all.x = TRUE)
```

```{r verify-counts, echo=TRUE}
# Row count = unique participants who have at least one row in session 1 or 2.
# If you expected more rows than nrow(wide), check: (1) knit cwd / wrong CSV file
# (fixed by root.dir in setup), (2) duplicate participant_id within a session
# (dedup keeps latest submitted_at), (3) participants only in session_num outside {1,2}.
pid_s12 <- unique(dt[session_num %in% c(1L, 2L), participant_id])
pid_other <- setdiff(unique(dt$participant_id), pid_s12)
if (length(pid_other)) {
  message(
    "Participants present in long data but with no session 1 or 2 row (excluded from wide): ",
    length(pid_other)
  )
  print(pid_other)
}
stopifnot(uniqueN(s1$participant_id) == nrow(s1))
stopifnot(uniqueN(s2$participant_id) == nrow(s2))
stopifnot(nrow(wide) == length(pid_s12))
stopifnot(setequal(wide$participant_id, pid_s12))
stopifnot(sum(is.na(wide$participant_age)) == 0L)
stopifnot(sum(is.na(wide$participant_gender)) == 0L)
```

---

## Derived design and outcome columns

- **`completed_both`:** `1` if both **`plank_s1`** and **`plank_s2`** are non-missing.
- **`crossover_order`:** From **`audio_s1`** — which condition was heard first (only defined when session 1 audio is present).
- **`first_session_is_high_tempo`:** Binary indicator for randomization arm.
- **`plank_high_tempo`** / **`plank_low_tempo`:** Plank duration under each condition, filled from whichever session had that **`audio_*`**.
- **`diff_high_minus_low`:** Paired contrast when both conditions observed; otherwise may be `NA`.
- **`elapsed_sessions_hours`:** `difftime(session 2, session 1, units = "hours")` in UTC, only when both timestamps exist and **`completed_both == 1`**; else **`NA`**.
- **`participant_*`:** Non-PII participant metadata merged from the participant export for later analysis without reopening raw participant files.

Finally, raw session submission timestamps are **removed** from the exported wide file so only **`elapsed_sessions_hours`** plus the participant-planned timing metadata remain.

```{r derive-wide}
wide[, completed_both := as.integer(!is.na(plank_s1) & !is.na(plank_s2))]

wide[, crossover_order := NA_character_]
wide[audio_s1 == "high_tempo", crossover_order := "high_tempo_first"]
wide[audio_s1 == "low_tempo", crossover_order := "low_tempo_first"]

wide[, first_session_is_high_tempo := NA_integer_]
wide[audio_s1 == "high_tempo", first_session_is_high_tempo := 1L]
wide[audio_s1 == "low_tempo", first_session_is_high_tempo := 0L]

wide[, plank_high_tempo := NA_real_]
wide[, plank_low_tempo := NA_real_]
wide[audio_s1 == "high_tempo", plank_high_tempo := plank_s1]
wide[audio_s1 == "low_tempo", plank_low_tempo := plank_s1]
wide[audio_s2 == "high_tempo", plank_high_tempo := plank_s2]
wide[audio_s2 == "low_tempo", plank_low_tempo := plank_s2]

wide[, diff_high_minus_low := plank_high_tempo - plank_low_tempo]

wide[, elapsed_sessions_hours := NA_real_]
ok_time <- !is.na(wide$submitted_at_s1) & !is.na(wide$submitted_at_s2) & (wide$completed_both == 1L)
wide[ok_time, elapsed_sessions_hours := as.numeric(difftime(submitted_at_s2, submitted_at_s1, units = "hours"))]

wide[, c("submitted_at_s1", "submitted_at_s2") := NULL]

core <- c(
  "participant_id", "completed_both", "crossover_order", "first_session_is_high_tempo",
  "audio_s1", "audio_s2", "plank_s1", "plank_s2",
  "plank_high_tempo", "plank_low_tempo", "diff_high_minus_low",
  "elapsed_sessions_hours"
)
participant_meta <- intersect(
  c(
    "participant_age", "participant_gender", "participant_locale",
    "participant_group_index", "participant_group_label",
    "participant_sessions_completed", "participant_registered_at",
    "participant_session1_planned_at", "participant_session2_planned_at"
  ),
  names(wide)
)
rest <- setdiff(names(wide), c(core, participant_meta))
setcolorder(wide, c(core, participant_meta, sort(rest)))
```

---

## Write outputs and codebook

The codebook is assembled from **`codebook_rows`**, deduplicated, and **restricted to column names that actually appear in `wide`** so orphan mappings (e.g. pre-task keys that never occurred for a session) are not shipped.

```{r write-files}
fwrite(wide, OUT_WIDE)

cb <- rbindlist(codebook_rows, fill = TRUE)
if (nrow(cb)) {
  cb <- unique(cb, by = c("column_name", "code_value"))
  cb <- cb[column_name %in% names(wide)]
  setorder(cb, column_name, code_value)
}
fwrite(cb, OUT_CODEBOOK)
```

---

## Run summary

```{r summary}
message("Wrote ", nrow(wide), " rows to ", OUT_WIDE)
message("Wrote ", nrow(cb), " codebook rows to ", OUT_CODEBOOK)

kable(
  wide[, .N, completed_both][order(completed_both)],
  caption = "Participants by completed_both (1 = both sessions)"
)
```

---

## Maintenance note

The standalone script **`process_plank_sessions_experimental.R`** mirrors this pipeline for non-interactive use (`Rscript ...`). If you change logic, update **either** the `.R` **or** this `.Rmd` and keep them in sync, or delete the `.R` and call `rmarkdown::render("process_plank_sessions_experimental.Rmd")` in automation instead.
