---
title: "Experiment analysis step 2: primary crossover analysis"
subtitle: "Paired descriptives, main figures, paired contrast, and fixed-effects model"
author: "W241 Plank Study"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  collapse = TRUE,
  fig.width = 7,
  fig.height = 4.4,
  dpi = 150
)

.analysis_dir <- {
  fn <- tryCatch(knitr::current_input(), error = function(e) "")
  if (is.null(fn) || !nzchar(fn)) {
    normalizePath(getwd(), winslash = "/")
  } else {
    dirname(normalizePath(fn, winslash = "/"))
  }
}
source(file.path(.analysis_dir, "00_analysis_setup.R"))
library(data.table)
library(ggplot2)
library(knitr)
library(sandwich)
library(lmtest)
```

## Purpose

This document estimates the primary within-subject effect of `high_tempo` versus `low_tempo` music on plank time-to-exhaustion using the canonical datasets created in Step 1.

The result sequence follows the report plan:

1. paired descriptives
2. paired mean-difference estimate
3. participant fixed-effects model with a period indicator

## Load analytic datasets

```{r load}
paired <- sanitize_dt_ascii(fread(derived_path("paired_analysis.csv")))
long_dt <- sanitize_dt_ascii(fread(derived_path("long_repeated_measures.csv")))
participant_flow <- sanitize_dt_ascii(fread(derived_path("participant_flow.csv")))

long_primary <- long_dt[in_primary_analysis == 1L]

stopifnot(nrow(paired) > 0L)
stopifnot(uniqueN(paired$participant_id) == nrow(paired))
stopifnot(nrow(long_primary) == 2L * nrow(paired))
stopifnot(uniqueN(long_primary$participant_id) == nrow(paired))
```

## Analytic sample

```{r sample-summary}
sample_tab <- data.table(
  quantity = c(
    "Paired analytic participants",
    "Repeated-measures rows used in FE model",
    "High-tempo first",
    "Low-tempo first"
  ),
  value = c(
    nrow(paired),
    nrow(long_primary),
    paired[, sum(crossover_order == "high_tempo_first")],
    paired[, sum(crossover_order == "low_tempo_first")]
  )
)

kable(sample_tab, caption = "Primary crossover sample")
```

## Descriptive summaries

```{r descriptives}
desc_tab <- data.table(
  quantity = c(
    "Mean high-tempo TTE",
    "Mean low-tempo TTE",
    "Mean paired difference (high - low)",
    "SD paired difference",
    "Median paired difference",
    "IQR paired difference",
    "Independent participants"
  ),
  value = c(
    mean(paired$high_tempo_tte),
    mean(paired$low_tempo_tte),
    mean(paired$diff_high_minus_low),
    sd(paired$diff_high_minus_low),
    median(paired$diff_high_minus_low),
    IQR(paired$diff_high_minus_low),
    nrow(paired)
  )
)

desc_tab[, value := round(value, 3)]
kable(desc_tab, caption = "Paired descriptive summaries")
```

## Compact metrics summary

```{r compact-metrics}
mean_high <- mean(paired$high_tempo_tte)
mean_low <- mean(paired$low_tempo_tte)
effect_mean <- mean(paired$diff_high_minus_low)
sd_diff <- sd(paired$diff_high_minus_low)
t_metrics <- t.test(paired$high_tempo_tte, paired$low_tempo_tte, paired = TRUE)
percent_improvement <- 100 * effect_mean / mean_low
completion_rate_registered <- 100 * participant_flow[, sum(completed_both_observed)] / nrow(participant_flow)

metrics_tab <- data.table(
  metric = c(
    "Mean treatment effect (high - low, seconds)",
    "Percentage improvement vs low tempo mean",
    "Standard deviation of the paired difference",
    "Paired t-test p-value",
    "Completion rate (completed both / registered)"
  ),
  value = c(
    sprintf("%.2f", effect_mean),
    sprintf("%.1f%%", percent_improvement),
    sprintf("%.2f", sd_diff),
    sprintf("%.4f", t_metrics$p.value),
    sprintf("%.1f%%", completion_rate_registered)
  )
)

kable(metrics_tab, caption = "Requested summary metrics")
```

## Exploratory figures

### Spaghetti plot

```{r fig-spaghetti}
plot_long <- melt(
  paired,
  id.vars = c("participant_id", "diff_high_minus_low"),
  measure.vars = c("low_tempo_tte", "high_tempo_tte"),
  variable.name = "condition",
  value.name = "seconds"
)

plot_long[, condition := factor(
  condition,
  levels = c("low_tempo_tte", "high_tempo_tte"),
  labels = c("Low tempo", "High tempo")
)]
plot_long[, favors_high := diff_high_minus_low > 0]

ggplot(plot_long, aes(x = condition, y = seconds, group = participant_id)) +
  geom_line(aes(color = favors_high), linewidth = 0.35, alpha = 0.75) +
  geom_point(aes(color = favors_high), size = 1.8, alpha = 0.9) +
  scale_color_manual(
    name = "High longer?",
    values = c(`TRUE` = "#1b7837", `FALSE` = "#c51b7d"),
    labels = c("FALSE" = "No", "TRUE" = "Yes")
  ) +
  labs(
    title = "Paired plank times by music condition",
    subtitle = paste0("n = ", uniqueN(plot_long$participant_id), " primary-sample participants"),
    x = NULL,
    y = "Plank duration (seconds)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

### Paired-difference distribution

```{r fig-diff}
ggplot(paired, aes(x = diff_high_minus_low)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 12,
    fill = "#3182bd",
    color = "white",
    linewidth = 0.2
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray30") +
  geom_vline(
    xintercept = mean(paired$diff_high_minus_low),
    color = "#e6550d",
    linewidth = 0.9
  ) +
  labs(
    title = "Distribution of paired differences",
    subtitle = "Orange = sample mean; dashed = no difference",
    x = "High tempo - low tempo (seconds)",
    y = "Density"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

## Model 1: raw paired contrast

```{r paired-contrast}
t_paired <- t.test(paired$high_tempo_tte, paired$low_tempo_tte, paired = TRUE)
effect_se <- sd(paired$diff_high_minus_low) / sqrt(nrow(paired))
effect_ci <- unname(t_paired$conf.int)
effect_dz <- effect_mean / sd(paired$diff_high_minus_low)

paired_results <- data.table(
  quantity = c(
    "Mean high-tempo TTE",
    "Mean low-tempo TTE",
    "Mean effect (high - low)",
    "95% CI lower",
    "95% CI upper",
    "Paired t statistic",
    "Paired p-value",
    "Standard error",
    "Paired effect size dz"
  ),
  value = c(
    mean(paired$high_tempo_tte),
    mean(paired$low_tempo_tte),
    effect_mean,
    effect_ci[1],
    effect_ci[2],
    unname(t_paired$statistic),
    t_paired$p.value,
    effect_se,
    effect_dz
  )
)

paired_results[, value := round(value, 4)]
kable(paired_results, caption = "Model 1. Raw paired contrast")
```

## Model 2: fixed-effects crossover regression

```{r fe-model}
long_primary[, participant_id := factor(participant_id)]
long_primary[, period_factor := factor(period)]

fe_fit <- lm(
  outcome_tte ~ high_tempo + period_factor + participant_id,
  data = long_primary
)

fe_vcov <- sandwich::vcovCL(fe_fit, cluster = ~ participant_id, type = "HC1")
fe_ct <- lmtest::coeftest(fe_fit, vcov. = fe_vcov)

beta_high <- unname(fe_ct["high_tempo", "Estimate"])
se_high <- unname(fe_ct["high_tempo", "Std. Error"])
ci_high <- beta_high + c(-1, 1) * 1.96 * se_high

beta_period2 <- unname(fe_ct["period_factor2", "Estimate"])
se_period2 <- unname(fe_ct["period_factor2", "Std. Error"])
ci_period2 <- beta_period2 + c(-1, 1) * 1.96 * se_period2

fe_results <- data.table(
  term = c("High-tempo indicator", "Period 2 indicator"),
  estimate = c(beta_high, beta_period2),
  std_error = c(se_high, se_period2),
  conf_low = c(ci_high[1], ci_period2[1]),
  conf_high = c(ci_high[2], ci_period2[2]),
  p_value = c(
    unname(fe_ct["high_tempo", "Pr(>|t|)"]),
    unname(fe_ct["period_factor2", "Pr(>|t|)"])
  )
)

fe_results[, `:=`(
  estimate = round(estimate, 4),
  std_error = round(std_error, 4),
  conf_low = round(conf_low, 4),
  conf_high = round(conf_high, 4),
  p_value = round(p_value, 4)
)]

kable(fe_results, caption = "Model 2. Fixed-effects crossover regression with clustered SEs")
```

## Main-effect comparison figure

```{r fig-main-effects, fig.height=3.8}
coef_plot <- rbind(
  data.table(
    model = "Paired contrast",
    estimate = effect_mean,
    conf_low = effect_ci[1],
    conf_high = effect_ci[2]
  ),
  data.table(
    model = "FE model",
    estimate = beta_high,
    conf_low = ci_high[1],
    conf_high = ci_high[2]
  )
)
coef_plot[, model := factor(model, levels = model)]

ggplot(coef_plot, aes(x = model, y = estimate, ymin = conf_low, ymax = conf_high)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_pointrange(color = "#3182bd", linewidth = 0.55, size = 0.75) +
  labs(
    title = "Primary effect estimates",
    subtitle = "Effect on plank time in seconds (high tempo - low tempo)",
    x = NULL,
    y = "Estimated effect (seconds)"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

## Verification

```{r verify}
stopifnot(nrow(paired) == 38L)
stopifnot(uniqueN(long_primary$participant_id) == nrow(paired))
stopifnot(all(long_primary$in_primary_analysis == 1L))

message("Primary analytic participants: ", nrow(paired))
message("Repeated-measures rows: ", nrow(long_primary))
message("Paired mean difference (high - low): ", round(effect_mean, 3))
message("FE high-tempo coefficient: ", round(beta_high, 3))
```

