---
title: "Experiment analysis step 3: period and order diagnostics"
subtitle: "Session-period checks for the primary crossover sample"
author: "W241 Plank Study"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  collapse = TRUE,
  fig.width = 7,
  fig.height = 4.3,
  dpi = 150
)

.analysis_dir <- {
  fn <- tryCatch(knitr::current_input(), error = function(e) "")
  if (is.null(fn) || !nzchar(fn)) {
    normalizePath(getwd(), winslash = "/")
  } else {
    dirname(normalizePath(fn, winslash = "/"))
  }
}
source(file.path(.analysis_dir, "00_analysis_setup.R"))
library(data.table)
library(ggplot2)
library(knitr)
library(sandwich)
library(lmtest)
```

## Purpose

This document checks whether broad period effects or order patterns appear large enough to threaten interpretation of the main crossover estimate.

These diagnostics are descriptive first. They do not replace the primary model; they help explain whether session order or repeated exposure appears material and whether the period term in the fixed-effects model is doing useful work.

## Load analytic datasets

```{r load}
paired <- sanitize_dt_ascii(fread(derived_path("paired_analysis.csv")))
long_dt <- sanitize_dt_ascii(fread(derived_path("long_repeated_measures.csv")))

long_primary <- long_dt[in_primary_analysis == 1L]

stopifnot(uniqueN(long_primary$participant_id) == nrow(paired))
stopifnot(nrow(long_primary) == 2L * nrow(paired))
```

## Period summaries

```{r period-summary}
period_summary <- long_primary[
  ,
  .(
    n = .N,
    mean_tte = mean(outcome_tte),
    sd_tte = sd(outcome_tte),
    median_tte = median(outcome_tte),
    iqr_tte = IQR(outcome_tte)
  ),
  by = period
][order(period)]

period_summary[, `:=`(
  mean_tte = round(mean_tte, 3),
  sd_tte = round(sd_tte, 3),
  median_tte = round(median_tte, 3),
  iqr_tte = round(iqr_tte, 3)
)]

kable(period_summary, caption = "Session 1 vs Session 2 summaries")
```

## Period-by-order summaries

```{r order-period-summary}
order_period_summary <- long_primary[
  ,
  .(
    n = .N,
    mean_tte = mean(outcome_tte),
    sd_tte = sd(outcome_tte),
    mean_rpe = mean(post_rpe, na.rm = TRUE)
  ),
  by = .(crossover_order, period, treatment)
][order(crossover_order, period)]

order_period_summary[, `:=`(
  mean_tte = round(mean_tte, 3),
  sd_tte = round(sd_tte, 3),
  mean_rpe = round(mean_rpe, 3)
)]

kable(order_period_summary, caption = "Period-by-order diagnostic summary")
```

## Session-period figures

### Session 1 vs Session 2 boxplot

```{r fig-period-box}
long_primary[, period_label := factor(period, levels = c(1, 2), labels = c("Session 1", "Session 2"))]

ggplot(long_primary, aes(x = period_label, y = outcome_tte, fill = period_label)) +
  geom_violin(alpha = 0.35, color = NA, trim = FALSE) +
  geom_boxplot(width = 0.12, fill = "white", outlier.shape = NA) +
  geom_jitter(width = 0.08, height = 0, alpha = 0.55, size = 1.6) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(
    title = "Outcome distributions by session period",
    subtitle = paste0("Primary analytic sample, n = ", uniqueN(long_primary$participant_id)),
    x = NULL,
    y = "Plank duration (seconds)"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

### Order-arm trajectories across periods

```{r fig-order-period}
order_plot <- long_primary[
  ,
  .(mean_tte = mean(outcome_tte)),
  by = .(crossover_order, period)
]

ggplot(order_plot, aes(x = period, y = mean_tte, color = crossover_order, group = crossover_order)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.1) +
  scale_x_continuous(breaks = c(1, 2), labels = c("Session 1", "Session 2")) +
  labs(
    title = "Average plank time by order arm across periods",
    subtitle = "Checks for sequence or learning patterns",
    x = NULL,
    y = "Mean plank duration (seconds)",
    color = "Order arm"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

## Diagnostic models

### Broad period check

```{r period-check}
period_check <- t.test(
  long_primary[period == 2, outcome_tte],
  long_primary[period == 1, outcome_tte],
  paired = FALSE
)

period_check_tab <- data.table(
  quantity = c(
    "Mean Session 2",
    "Mean Session 1",
    "Difference (Session 2 - Session 1)",
    "95% CI lower",
    "95% CI upper",
    "p-value"
  ),
  value = c(
    mean(long_primary[period == 2, outcome_tte]),
    mean(long_primary[period == 1, outcome_tte]),
    diff(rev(period_check$estimate)),
    period_check$conf.int[1],
    period_check$conf.int[2],
    period_check$p.value
  )
)

period_check_tab[, value := round(value, 4)]
kable(period_check_tab, caption = "Broad period comparison")
```

### Sequence diagnostic regression

```{r sequence-regression}
long_primary[, crossover_order := factor(crossover_order)]
long_primary[, period_factor := factor(period)]

diag_fit <- lm(
  outcome_tte ~ period_factor * crossover_order,
  data = long_primary
)

diag_vcov <- sandwich::vcovCL(diag_fit, cluster = ~ participant_id, type = "HC1")
diag_ct <- lmtest::coeftest(diag_fit, vcov. = diag_vcov)

diag_terms <- intersect(
  c("period_factor2", "crossover_orderlow_tempo_first", "period_factor2:crossover_orderlow_tempo_first"),
  rownames(diag_ct)
)

diag_results <- data.table(
  term = diag_terms,
  estimate = diag_ct[diag_terms, "Estimate"],
  std_error = diag_ct[diag_terms, "Std. Error"],
  p_value = diag_ct[diag_terms, "Pr(>|t|)"]
)
diag_results[, `:=`(
  estimate = round(estimate, 4),
  std_error = round(std_error, 4),
  p_value = round(p_value, 4)
)]

kable(diag_results, caption = "Sequence diagnostic regression")
```

## Interpretation notes

```{r interpretation}
period_effect <- round(diag_ct["period_factor2", "Estimate"], 3)
interaction_effect <- round(diag_ct["period_factor2:crossover_orderlow_tempo_first", "Estimate"], 3)

message("Estimated broad period effect (Session 2 vs Session 1): ", period_effect, " seconds")
message("Estimated period-by-order interaction: ", interaction_effect, " seconds")
```

## Verification

```{r verify}
stopifnot(uniqueN(long_primary$participant_id) == 38L)
stopifnot(all(long_primary$period %in% c(1L, 2L)))
stopifnot(nrow(order_period_summary) == 4L)

message("Diagnostic sample participants: ", uniqueN(long_primary$participant_id))
message("Diagnostic sample rows: ", nrow(long_primary))
```

