---
title: "Experiment analysis step 5: precision comparison"
subtitle: "Realized precision for crossover and first-period benchmark designs"
author: "W241 Plank Study"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  collapse = TRUE,
  fig.width = 7,
  fig.height = 4.3,
  dpi = 150
)

.analysis_dir <- {
  fn <- tryCatch(knitr::current_input(), error = function(e) "")
  if (is.null(fn) || !nzchar(fn)) {
    normalizePath(getwd(), winslash = "/")
  } else {
    dirname(normalizePath(fn, winslash = "/"))
  }
}
source(file.path(.analysis_dir, "00_analysis_setup.R"))
library(data.table)
library(ggplot2)
library(knitr)
```

## Purpose

This document compares the realized precision of:

- the primary crossover design
- the first-period-only parallel benchmark

The comparison is based on standard errors, 95% CI widths, minimum detectable effects, and the benchmark sample size required to match crossover precision. This is more informative than retrospective observed power.

## Load datasets

```{r load}
paired <- sanitize_dt_ascii(fread(derived_path("paired_analysis.csv")))
long_dt <- sanitize_dt_ascii(fread(derived_path("long_repeated_measures.csv")))

bench <- long_dt[
  in_first_period_benchmark == 1L & period == 1L
]

stopifnot(nrow(paired) == 38L)
stopifnot(nrow(bench) == 46L)
```

## Precision calculations

```{r calculations}
alpha_two_sided <- 0.05
target_power <- 0.80
z_alpha <- qnorm(1 - alpha_two_sided / 2)
z_power <- qnorm(target_power)
mde_multiplier <- z_alpha + z_power

# Crossover / paired design
n_pair <- nrow(paired)
sd_diff <- sd(paired$diff_high_minus_low)
se_pair <- sd_diff / sqrt(n_pair)
ci_width_pair <- 2 * z_alpha * se_pair
mde_pair <- mde_multiplier * se_pair

# First-period parallel benchmark
n_high <- bench[high_tempo == 1L, .N]
n_low <- bench[high_tempo == 0L, .N]
sd_high <- bench[high_tempo == 1L, sd(outcome_tte)]
sd_low <- bench[high_tempo == 0L, sd(outcome_tte)]
se_parallel <- sqrt((sd_high^2 / n_high) + (sd_low^2 / n_low))
ci_width_parallel <- 2 * z_alpha * se_parallel
mde_parallel <- mde_multiplier * se_parallel

sd_pooled <- sqrt(
  (((n_high - 1) * sd_high^2) + ((n_low - 1) * sd_low^2)) /
    (n_high + n_low - 2)
)

# Equal-allocation parallel design needed to match the crossover SE
n_per_arm_to_match <- ceiling((2 * sd_pooled^2) / (se_pair^2))
total_n_to_match <- 2 * n_per_arm_to_match

precision_tab <- data.table(
  design = c("Primary crossover", "First-period benchmark"),
  independent_n = c(n_pair, n_high + n_low),
  observations = c(2 * n_pair, n_high + n_low),
  relevant_sd = c(sd_diff, sd_pooled),
  standard_error = c(se_pair, se_parallel),
  ci_width_95 = c(ci_width_pair, ci_width_parallel),
  mde_80_power = c(mde_pair, mde_parallel)
)

precision_tab[, `:=`(
  relevant_sd = round(relevant_sd, 4),
  standard_error = round(standard_error, 4),
  ci_width_95 = round(ci_width_95, 4),
  mde_80_power = round(mde_80_power, 4)
)]
```

## Precision table

```{r precision-table}
kable(precision_tab, caption = "Realized precision by design")
```

```{r matching-n}
matching_tab <- data.table(
  quantity = c(
    "Observed benchmark total N",
    "Required benchmark total N to match crossover SE",
    "Required benchmark N per arm to match crossover SE",
    "Extra independent participants needed"
  ),
  value = c(
    n_high + n_low,
    total_n_to_match,
    n_per_arm_to_match,
    total_n_to_match - (n_high + n_low)
  )
)

kable(matching_tab, caption = "Parallel-design sample size needed to match crossover precision")
```

## Precision comparison figure

```{r fig-precision}
plot_precision <- melt(
  precision_tab[, .(design, standard_error, ci_width_95, mde_80_power)],
  id.vars = "design",
  variable.name = "metric",
  value.name = "value"
)

plot_precision[, metric := factor(
  metric,
  levels = c("standard_error", "ci_width_95", "mde_80_power"),
  labels = c("Standard error", "95% CI width", "MDE (80% power)")
)]

ggplot(plot_precision, aes(x = metric, y = value, fill = design)) +
  geom_col(position = "dodge") +
  labs(
    title = "Realized precision comparison",
    subtitle = "Smaller values indicate tighter design precision",
    x = NULL,
    y = "Seconds",
    fill = "Design"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

## Interpretation notes

```{r interpretation}
precision_ratio <- se_parallel / se_pair
ci_ratio <- ci_width_parallel / ci_width_pair

interpretation_tab <- data.table(
  quantity = c(
    "Parallel SE / crossover SE",
    "Parallel CI width / crossover CI width"
  ),
  value = round(c(precision_ratio, ci_ratio), 4)
)

kable(interpretation_tab, caption = "Relative precision loss in the benchmark design")
```

## Verification

```{r verify}
stopifnot(n_pair == 38L)
stopifnot((n_high + n_low) == 46L)
stopifnot(se_pair > 0)
stopifnot(se_parallel > 0)
stopifnot(total_n_to_match >= (n_high + n_low))

message("Crossover SE: ", round(se_pair, 3))
message("Benchmark SE: ", round(se_parallel, 3))
message("Required parallel total N to match crossover SE: ", total_n_to_match)
```

