---
title: "Experiment analysis step 6: secondary outcomes and HTE"
subtitle: "Exploratory secondary outcomes and heterogeneous treatment effects"
author: "W241 Plank Study"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  collapse = TRUE
)

.analysis_dir <- {
  fn <- tryCatch(knitr::current_input(), error = function(e) "")
  if (is.null(fn) || !nzchar(fn)) {
    normalizePath(getwd(), winslash = "/")
  } else {
    dirname(normalizePath(fn, winslash = "/"))
  }
}
source(file.path(.analysis_dir, "00_analysis_setup.R"))
library(data.table)
library(knitr)
library(sandwich)
library(lmtest)
```

## Purpose

This document treats secondary outcomes and subgroup effects as exploratory.

- Secondary outcomes: `RPE`, `motivation`, and `energy`
- HTE targets: `age`, `gender`, and baseline `activity level` as a proxy for fitness

These results are reported after the main crossover estimate and interpreted cautiously.

## Load analytic datasets

```{r load}
paired <- sanitize_dt_ascii(fread(derived_path("paired_analysis.csv")))
long_dt <- sanitize_dt_ascii(fread(derived_path("long_repeated_measures.csv")))
long_primary <- long_dt[in_primary_analysis == 1L]

stopifnot(nrow(paired) == 38L)
stopifnot(uniqueN(long_primary$participant_id) == 38L)
```

## Secondary outcomes: paired contrasts

```{r paired-secondary}
secondary_specs <- list(
  list(label = "RPE", high = "rpe_high_tempo", low = "rpe_low_tempo"),
  list(label = "Motivation", high = "motivation_high_tempo", low = "motivation_low_tempo"),
  list(label = "Energy", high = "energy_high_tempo", low = "energy_low_tempo")
)

secondary_rows <- lapply(secondary_specs, function(spec) {
  hi <- paired[[spec$high]]
  lo <- paired[[spec$low]]
  tt <- t.test(hi, lo, paired = TRUE)
  data.table(
    outcome = spec$label,
    mean_high = mean(hi, na.rm = TRUE),
    mean_low = mean(lo, na.rm = TRUE),
    mean_diff_high_minus_low = mean(hi - lo, na.rm = TRUE),
    conf_low = tt$conf.int[1],
    conf_high = tt$conf.int[2],
    p_value = tt$p.value
  )
})

secondary_paired <- rbindlist(secondary_rows)
secondary_paired[, `:=`(
  mean_high = round(mean_high, 4),
  mean_low = round(mean_low, 4),
  mean_diff_high_minus_low = round(mean_diff_high_minus_low, 4),
  conf_low = round(conf_low, 4),
  conf_high = round(conf_high, 4),
  p_value = round(p_value, 4)
)]

kable(secondary_paired, caption = "Exploratory paired contrasts for secondary outcomes")
```

## Secondary outcomes: fixed-effects versions

```{r fe-secondary}
run_fe_secondary <- function(data, outcome_var, label) {
  tmp <- copy(data)
  tmp[["participant_id"]] <- factor(tmp[["participant_id"]])
  tmp[["period_factor"]] <- factor(tmp[["period"]])

  fit <- lm(
    reformulate(c("high_tempo", "period_factor", "participant_id"), response = outcome_var),
    data = tmp
  )
  vc <- sandwich::vcovCL(fit, cluster = ~ participant_id, type = "HC1")
  ct <- lmtest::coeftest(fit, vcov. = vc)

  data.table(
    outcome = label,
    estimate = unname(ct["high_tempo", "Estimate"]),
    std_error = unname(ct["high_tempo", "Std. Error"]),
    p_value = unname(ct["high_tempo", "Pr(>|t|)"])
  )
}

secondary_fe <- rbindlist(list(
  run_fe_secondary(long_primary, "post_rpe", "RPE"),
  run_fe_secondary(long_primary, "pre_motivation", "Motivation"),
  run_fe_secondary(long_primary, "pre_energy", "Energy")
))

secondary_fe[, `:=`(
  estimate = round(estimate, 4),
  std_error = round(std_error, 4),
  p_value = round(p_value, 4)
)]

kable(secondary_fe, caption = "Exploratory fixed-effects treatment coefficients for secondary outcomes")
```

## HTE setup

We estimate treatment interactions on the primary TTE outcome. Because subgroup results are exploratory, we report Bonferroni-adjusted p-values across the three planned subgroup tests.

```{r hte-setup}
hte_data <- copy(long_primary)
hte_data[, participant_id := factor(participant_id)]
hte_data[, period_factor := factor(period)]
hte_data[, age_centered := participant_age - mean(participant_age, na.rm = TRUE)]
hte_data[, male_indicator := as.integer(participant_gender == "male")]
hte_data[, activity_level_ord := as.numeric(activity_level_baseline)]
```

## HTE models

```{r hte-models}
run_hte <- function(data, subgroup_var, label) {
  fit <- lm(
    reformulate(
      c("high_tempo", subgroup_var, paste0("high_tempo:", subgroup_var), "period_factor", "participant_id"),
      response = "outcome_tte"
    ),
    data = data
  )
  vc <- sandwich::vcovCL(fit, cluster = ~ participant_id, type = "HC1")
  ct <- lmtest::coeftest(fit, vcov. = vc)
  term <- paste0("high_tempo:", subgroup_var)

  data.table(
    subgroup = label,
    interaction_term = term,
    estimate = unname(ct[term, "Estimate"]),
    std_error = unname(ct[term, "Std. Error"]),
    p_value = unname(ct[term, "Pr(>|t|)"])
  )
}

hte_results <- rbindlist(list(
  run_hte(hte_data, "age_centered", "Age"),
  run_hte(hte_data, "male_indicator", "Gender (male vs not male)"),
  run_hte(hte_data, "activity_level_ord", "Activity level / fitness proxy")
))

hte_results[, p_value_bonferroni := p.adjust(p_value, method = "bonferroni")]
hte_results[, `:=`(
  estimate = round(estimate, 4),
  std_error = round(std_error, 4),
  p_value = round(p_value, 4),
  p_value_bonferroni = round(p_value_bonferroni, 4)
)]

kable(hte_results, caption = "Exploratory HTE interaction results with Bonferroni correction")
```

## Verification

```{r verify}
stopifnot(nrow(secondary_paired) == 3L)
stopifnot(nrow(secondary_fe) == 3L)
stopifnot(nrow(hte_results) == 3L)

message("Secondary outcomes analyzed: ", paste(secondary_paired$outcome, collapse = ", "))
message("HTE tests analyzed: ", paste(hte_results$subgroup, collapse = ", "))
```

