---
title: "W241 Plank Study Experimental Report"
subtitle: "Crossover and benchmark analysis"
author: "W241 Plank Study"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    latex_engine: pdflatex
---

```{r setup, include=FALSE}
options(knitr.duplicate.label = "allow")

library(knitr)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  fig.width = 7,
  fig.height = 4.4,
  dpi = 150
)

.analysis_dir <- {
  fn <- tryCatch(knitr::current_input(), error = function(e) "")
  if (is.null(fn) || !nzchar(fn)) {
    normalizePath(getwd(), winslash = "/")
  } else {
    dirname(normalizePath(fn, winslash = "/"))
  }
}
source(file.path(.analysis_dir, "00_analysis_setup.R"))
library(data.table)
library(ggplot2)
library(knitr)
library(sandwich)
library(lmtest)

strip_yaml_and_setup <- function(path) {
  lines <- readLines(path, warn = FALSE)

  if (length(lines) >= 1L && trimws(lines[1L]) == "---") {
    end_yaml <- which(trimws(lines[-1L]) == "---")[1L] + 1L
    lines <- lines[(end_yaml + 1L):length(lines)]
  }

  setup_start <- grep("^```\\{r setup", lines)[1L]
  if (!is.na(setup_start)) {
    setup_end <- setup_start + which(lines[setup_start:length(lines)] == "```")[1L] - 1L
    lines <- lines[-seq.int(setup_start, setup_end)]
  }

  purpose_start <- match("## Purpose", lines)
  if (!is.na(purpose_start)) {
    next_heading <- grep("^## ", lines[(purpose_start + 1L):length(lines)])[1L]
    if (is.na(next_heading)) {
      lines <- lines[seq_len(purpose_start - 1L)]
    } else {
      purpose_end <- purpose_start + next_heading - 1L
      lines <- lines[-seq.int(purpose_start, purpose_end - 1L)]
    }
  }

  heading_only_patterns <- c("^## Load ", "^## Verification$")
  for (pat in heading_only_patterns) {
    hit <- grep(pat, lines)
    if (length(hit)) lines <- lines[-hit]
  }

  lines
}

include_section <- function(filename) {
  body <- strip_yaml_and_setup(file.path(.analysis_dir, filename))
  cat(knit_child(text = body, envir = environment(), quiet = TRUE), sep = "\n")
}
```

## Report Structure

This report follows the order in the experiment analysis plan:

1. participant flow, attrition, and exclusions
2. primary crossover effect
3. period and order diagnostics
4. first-period benchmark
5. realized precision comparison
6. secondary outcomes and HTE
7. randomization inference

## Study Overview

The primary estimand is the **within-person effect of high-tempo versus low-tempo music on plank time-to-exhaustion**, measured in seconds. The main estimate uses the randomized crossover design, so each participant serves as their own control and stable participant-level differences are absorbed by construction.

The first-period benchmark is supplementary. It asks what the treatment contrast would have looked like if the study had been analyzed as a standard parallel-arm design using only session 1. That benchmark is included to clarify the design choice, not to replace the crossover estimate.

## Participant Flow And Analysis Populations

The report begins with the full registered participant roster, then narrows to observed outcomes and protocol-valid analytic samples. This keeps attrition, missingness, and protocol exclusions visible before any inferential models are introduced.

```{r section-01, results='asis'}
include_section("01_analysis_populations.Rmd")
```

## Primary Crossover Estimate

With the populations fixed, the next step is the primary within-subject treatment estimate. The emphasis here is the design-consistent sequence: paired descriptives first, then the paired mean difference, then a participant fixed-effects model without period adjustment, and finally the fixed-effects crossover model with a period adjustment.

```{r section-02, results='asis'}
include_section("02_primary_crossover_analysis.Rmd")
```

## Period And Order Diagnostics

After the main estimate, the report checks whether broad period effects or order patterns appear large enough to complicate interpretation. These diagnostics are supportive rather than alternative estimators.

```{r section-03, results='asis'}
include_section("03_period_order_diagnostics.Rmd")
```

## First-Period Benchmark

Only after the primary crossover evidence and the period/order checks do we turn to the first-period benchmark. This keeps the benchmark in its intended role: a secondary comparison that illustrates what a parallel-arm analysis would have looked like before carryover could matter.

```{r section-04, results='asis'}
include_section("04_first_period_benchmark.Rmd")
```

## Results At A Glance

The key treatment-effect results are consolidated here so the reader can compare the primary crossover estimate, the supplementary benchmark, and the design-based inference in one place.

```{r results-at-a-glance}
paired_master <- sanitize_dt_ascii(fread(derived_path("paired_analysis.csv")))
long_master <- sanitize_dt_ascii(fread(derived_path("long_repeated_measures.csv")))
long_primary_master <- copy(long_master[in_primary_analysis == 1L])
bench_master <- copy(long_master[in_first_period_benchmark == 1L & period == 1L])

t_paired_master <- t.test(paired_master$high_tempo_tte, paired_master$low_tempo_tte, paired = TRUE)
paired_effect_master <- mean(paired_master$diff_high_minus_low)
paired_ci_master <- unname(t_paired_master$conf.int)

long_primary_master[["participant_id"]] <- factor(long_primary_master[["participant_id"]])
fe_noperiod_master <- lm(outcome_tte ~ high_tempo + participant_id, data = long_primary_master)
fe_noperiod_master_vcov <- sandwich::vcovCL(fe_noperiod_master, cluster = ~ participant_id, type = "HC1")
fe_noperiod_master_ct <- lmtest::coeftest(fe_noperiod_master, vcov. = fe_noperiod_master_vcov)
fe_noperiod_effect_master <- unname(fe_noperiod_master_ct["high_tempo", "Estimate"])
fe_noperiod_p_master <- unname(fe_noperiod_master_ct["high_tempo", "Pr(>|t|)"])

long_primary_master[["period_factor"]] <- factor(long_primary_master[["period"]])
fe_master <- lm(outcome_tte ~ high_tempo + period_factor + participant_id, data = long_primary_master)
fe_master_vcov <- sandwich::vcovCL(fe_master, cluster = ~ participant_id, type = "HC1")
fe_master_ct <- lmtest::coeftest(fe_master, vcov. = fe_master_vcov)
fe_effect_master <- unname(fe_master_ct["high_tempo", "Estimate"])
fe_p_master <- unname(fe_master_ct["high_tempo", "Pr(>|t|)"])

t_bench_master <- t.test(outcome_tte ~ high_tempo, data = bench_master)
bench_effect_master <- bench_master[high_tempo == 1L, mean(outcome_tte)] - bench_master[high_tempo == 0L, mean(outcome_tte)]
bench_ci_master <- unname(t_bench_master$conf.int)

set.seed(241)
ri_dist_master <- replicate(
  10000L,
  mean(paired_master$diff_high_minus_low * sample(c(-1, 1), nrow(paired_master), replace = TRUE))
)
ri_p_master <- mean(abs(ri_dist_master) >= abs(paired_effect_master))

main_results_tab <- data.table(
  statistic = c(
    "Paired analytic participants",
    "Mean low-tempo TTE",
    "Mean high-tempo TTE",
    "Paired effect (high - low)",
    "Paired 95% CI",
    "Paired p-value",
    "FE without-period estimate",
    "FE without-period p-value",
    "FE with-period estimate",
    "FE with-period p-value",
    "RI two-sided p-value"
  ),
  value = c(
    as.character(nrow(paired_master)),
    sprintf("%.2f", mean(paired_master$low_tempo_tte)),
    sprintf("%.2f", mean(paired_master$high_tempo_tte)),
    sprintf("%.2f", paired_effect_master),
    sprintf("[%.2f, %.2f]", paired_ci_master[1], paired_ci_master[2]),
    sprintf("%.4f", t_paired_master$p.value),
    sprintf("%.2f", fe_noperiod_effect_master),
    sprintf("%.4f", fe_noperiod_p_master),
    sprintf("%.2f", fe_effect_master),
    sprintf("%.4f", fe_p_master),
    sprintf("%.4f", ri_p_master)
  )
)

kable(main_results_tab, caption = "Compact main-results summary")
```

```{r fig-effect-comparison, fig.height=3.8}
effect_compare <- data.table(
  analysis = c("Primary crossover", "First-period benchmark"),
  estimate = c(paired_effect_master, bench_effect_master),
  conf_low = c(paired_ci_master[1], bench_ci_master[1]),
  conf_high = c(paired_ci_master[2], bench_ci_master[2])
)
effect_compare[, analysis := factor(analysis, levels = analysis)]

ggplot(effect_compare, aes(x = analysis, y = estimate, ymin = conf_low, ymax = conf_high)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_pointrange(color = "#3182bd", linewidth = 0.55, size = 0.75) +
  labs(
    title = "Treatment-effect comparison by design",
    subtitle = "Both estimates are shown in seconds: high tempo minus low tempo",
    x = NULL,
    y = "Estimated effect (seconds)"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

## Precision And Design-Efficiency Comparison

Once both the crossover and benchmark estimates are on the page, the precision comparison shows why the crossover design is more efficient for this question.

```{r section-05, results='asis'}
include_section("05_precision_power_comparison.Rmd")
```

## Exploratory Mechanisms And Heterogeneity

The next section keeps the mechanism-related outcomes and subgroup patterns clearly downstream of the primary estimate. These are exploratory checks rather than headline findings.

```{r section-06, results='asis'}
include_section("06_secondary_outcomes_and_hte.Rmd")
```

## Design-Based Inference Check

The report closes the analytic sequence with randomization inference, which checks whether the observed paired effect is unusual under the sharp null implied by the randomization mechanism.

```{r section-07, results='asis'}
include_section("07_randomization_inference.Rmd")
```

