---
title: "Final Experimental Report"
subtitle: "Beat the Burn: The Effect of Music Tempo on Isometric Plank Endurance"
author: "By: Ryan Castillo, Thomas Dobbs, Karan Patel, Carlos Schrupp"
date: "MIDS 241, 2026-04-15"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: false
    latex_engine: pdflatex
---

```{r setup, include=FALSE}
options(knitr.duplicate.label = "allow")

library(knitr)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  fig.width = 7,
  fig.height = 4.4,
  dpi = 150
)

.analysis_dir <- {
  fn <- tryCatch(knitr::current_input(), error = function(e) "")
  if (is.null(fn) || !nzchar(fn)) {
    normalizePath(getwd(), winslash = "/")
  } else {
    dirname(normalizePath(fn, winslash = "/"))
  }
}
source(file.path(.analysis_dir, "00_analysis_setup.R"))
library(data.table)
library(ggplot2)
library(sandwich)
library(lmtest)

strip_sections <- function(lines, heading_titles) {
  for (title in heading_titles) {
    start <- grep(paste0("^## ", title, "$"), lines)[1L]
    if (is.na(start)) next
    next_heading <- grep("^## ", lines[(start + 1L):length(lines)])[1L]
    if (is.na(next_heading)) {
      lines <- lines[seq_len(start - 1L)]
    } else {
      end <- start + next_heading - 2L
      lines <- lines[-seq.int(start, end)]
    }
  }
  lines
}

strip_yaml_and_setup <- function(path, drop_sections = character()) {
  lines <- readLines(path, warn = FALSE)

  if (length(lines) >= 1L && trimws(lines[1L]) == "---") {
    end_yaml <- which(trimws(lines[-1L]) == "---")[1L] + 1L
    lines <- lines[(end_yaml + 1L):length(lines)]
  }

  setup_start <- grep("^```\\{r setup", lines)[1L]
  if (!is.na(setup_start)) {
    setup_end <- setup_start + which(lines[setup_start:length(lines)] == "```")[1L] - 1L
    lines <- lines[-seq.int(setup_start, setup_end)]
  }

  purpose_start <- match("## Purpose", lines)
  if (!is.na(purpose_start)) {
    next_heading <- grep("^## ", lines[(purpose_start + 1L):length(lines)])[1L]
    if (is.na(next_heading)) {
      lines <- lines[seq_len(purpose_start - 1L)]
    } else {
      purpose_end <- purpose_start + next_heading - 1L
      lines <- lines[-seq.int(purpose_start, purpose_end - 1L)]
    }
  }

  heading_only_patterns <- c("^## Load ", "^## Verification$")
  for (pat in heading_only_patterns) {
    hit <- grep(pat, lines)
    if (length(hit)) lines <- lines[-hit]
  }

  if (length(drop_sections)) {
    lines <- strip_sections(lines, drop_sections)
  }

  lines
}

include_section <- function(filename, drop_sections = character()) {
  body <- strip_yaml_and_setup(file.path(.analysis_dir, filename), drop_sections = drop_sections)
  cat(knit_child(text = body, envir = environment(), quiet = TRUE), sep = "\n")
}

render_stargazer_table <- function(stem, caption = NULL) {
  if (!is.null(caption) && nzchar(caption)) {
    cat(caption, "\n\n")
  }
  if (knitr::is_latex_output()) {
    path <- file.path(.analysis_dir, "output", paste0(stem, ".tex"))
    lines <- readLines(path, warn = FALSE)
    cat(paste(lines, collapse = "\n"))
    cat("\n\n")
  } else {
    path <- file.path(.analysis_dir, "output", paste0(stem, ".txt"))
    lines <- readLines(path, warn = FALSE)
    cat("```text\n")
    cat(paste(lines, collapse = "\n"))
    cat("\n```\n\n")
  }
}

flow <- sanitize_dt_ascii(fread(derived_path("participant_flow.csv")))
paired <- sanitize_dt_ascii(fread(derived_path("paired_analysis.csv")))
long_dt <- sanitize_dt_ascii(fread(derived_path("long_repeated_measures.csv")))

long_primary <- copy(long_dt[in_primary_analysis == 1L])
bench <- copy(long_dt[in_first_period_benchmark == 1L & period == 1L])

long_primary[, participant_id := factor(participant_id)]
fe_noperiod <- lm(outcome_tte ~ high_tempo + participant_id, data = long_primary)
fe_noperiod_vcov <- sandwich::vcovCL(fe_noperiod, cluster = ~ participant_id, type = "HC1")
fe_noperiod_ct <- lmtest::coeftest(fe_noperiod, vcov. = fe_noperiod_vcov)

long_primary[, period_factor := factor(period)]
fe_period <- lm(outcome_tte ~ high_tempo + period_factor + participant_id, data = long_primary)
fe_period_vcov <- sandwich::vcovCL(fe_period, cluster = ~ participant_id, type = "HC1")
fe_period_ct <- lmtest::coeftest(fe_period, vcov. = fe_period_vcov)

t_paired <- t.test(paired$high_tempo_tte, paired$low_tempo_tte, paired = TRUE)
effect_mean <- mean(paired$diff_high_minus_low)
effect_ci <- unname(t_paired$conf.int)
percent_improvement <- 100 * effect_mean / mean(paired$low_tempo_tte)

bench_t <- t.test(outcome_tte ~ high_tempo, data = bench)
bench_effect <- bench[high_tempo == 1L, mean(outcome_tte)] - bench[high_tempo == 0L, mean(outcome_tte)]
bench_ci <- unname(bench_t$conf.int)

sd_diff <- sd(paired$diff_high_minus_low)
se_pair <- sd_diff / sqrt(nrow(paired))
n_high <- bench[high_tempo == 1L, .N]
n_low <- bench[high_tempo == 0L, .N]
sd_high <- bench[high_tempo == 1L, sd(outcome_tte)]
sd_low <- bench[high_tempo == 0L, sd(outcome_tte)]
se_parallel <- sqrt((sd_high^2 / n_high) + (sd_low^2 / n_low))
precision_ratio <- se_parallel / se_pair

set.seed(241)
ri_dist <- replicate(
  10000L,
  mean(paired$diff_high_minus_low * sample(c(-1, 1), nrow(paired), replace = TRUE))
)
ri_two_sided <- mean(abs(ri_dist) >= abs(effect_mean))

sec_pairs <- list(
  RPE = c("rpe_high_tempo", "rpe_low_tempo"),
  Motivation = c("motivation_high_tempo", "motivation_low_tempo"),
  Energy = c("energy_high_tempo", "energy_low_tempo")
)
sec_diff <- function(nm) mean(paired[[sec_pairs[[nm]][1]]] - paired[[sec_pairs[[nm]][2]]], na.rm = TRUE)
sec_p <- function(nm) t.test(paired[[sec_pairs[[nm]][1]]], paired[[sec_pairs[[nm]][2]]], paired = TRUE)$p.value

hte <- copy(long_primary)
hte[, age_centered := participant_age - mean(participant_age, na.rm = TRUE)]
hte[, male_indicator := as.integer(participant_gender == "male")]
hte[, activity_level_ord := as.numeric(activity_level_baseline)]
hte[, plank_familiarity_ord := fcase(
  plank_frequency_baseline == 1L, 1L,
  plank_frequency_baseline %in% c(2L, 3L), 2L,
  plank_frequency_baseline == 4L, 3L,
  default = NA_integer_
)]
run_hte_p <- function(var) {
  dat <- copy(hte)
  dat[["participant_id"]] <- factor(dat[["participant_id"]])
  dat[["period_factor"]] <- factor(dat[["period"]])
  fit <- lm(
    reformulate(c("high_tempo", var, paste0("high_tempo:", var), "period_factor", "participant_id"),
      response = "outcome_tte"),
    data = dat
  )
  vc <- sandwich::vcovCL(fit, cluster = ~ participant_id, type = "HC1")
  ct <- lmtest::coeftest(fit, vcov. = vc)
  ct[paste0("high_tempo:", var), "Pr(>|t|)"]
}
hte_ps_adj <- p.adjust(c(
  run_hte_p("age_centered"),
  run_hte_p("male_indicator"),
  run_hte_p("activity_level_ord"),
  run_hte_p("plank_familiarity_ord")
), method = "bonferroni")

model_roadmap <- data.table(
  family = c(
    "P: Primary crossover",
    "P: Primary crossover",
    "P: Primary crossover",
    "D: Diagnostics",
    "D: Diagnostics",
    "B: Benchmark",
    "B: Benchmark",
    "S: Secondary outcomes",
    "S: Secondary outcomes",
    "H: Heterogeneity",
    "R: Randomization inference",
    "R: Randomization inference"
  ),
  id = c("P1", "P2", "P3", "D1", "D2", "B1", "B2", "S1-S3", "S4-S6", "H1-H4", "R1", "R2"),
  model = c(
    "Raw paired contrast",
    "Fixed-effects without period",
    "Fixed-effects with period",
    "Broad period comparison",
    "Sequence diagnostic regression",
    "First-period raw contrast",
    "First-period adjusted regression",
    "Paired secondary contrasts",
    "Fixed-effects secondary models",
    "Treatment-by-subgroup interaction models",
    "Bridge linear model",
    "Randomization inference"
  ),
  sample = c(
    "38 completed pairs",
    "76 repeated-measures rows",
    "76 repeated-measures rows",
    "76 repeated-measures rows",
    "76 repeated-measures rows",
    "46 Session 1 observations",
    "46 Session 1 observations",
    "38 completed pairs",
    "76 repeated-measures rows",
    "76 repeated-measures rows",
    "38 participant-level differences",
    "38 participant-level differences"
  ),
  purpose = c(
    "Direct within-person treatment contrast",
    "Linear-model bridge without period adjustment",
    "Main regression specification with period control",
    "Checks broad Session 2 drift",
    "Checks whether period patterns differ by randomized order",
    "Emulates a parallel-arm comparison using Session 1 only",
    "Adds baseline adjustment to the Session 1 benchmark",
    "Tests simple paired shifts in RPE, motivation, and energy",
    "Re-estimates secondary outcomes in the FE framework",
    "Tests whether treatment effects differ by pre-specified subgroup",
    "Writes the paired estimand in regression form",
    "Provides design-based sign-flip p-values under the sharp null"
  )
)
```

# I. Introduction

This report studies the within-person causal effect of listening to high-tempo versus low-tempo instrumental EDM music on plank time-to-volitional-exhaustion (TTE), measured in seconds. The causal quantity of interest is the difference between each participant's potential outcome under the high-tempo condition and their potential outcome under the low-tempo condition, averaged over the retained analytic sample. The primary estimand is therefore the average treatment effect inside the randomized crossover design: for each participant, we observe plank duration under each condition, take the within-person difference, and average it across participants.

An observational comparison would not identify that estimand. People who naturally prefer fast music may also differ systematically in baseline strength, training history, motivation, pain tolerance, or willingness to push through discomfort. Those selection effects would contaminate any simple comparison of high-tempo listeners to low-tempo listeners. The experiment was therefore designed to randomize treatment order and let each participant serve as their own control, sharply reducing stable between-person confounding.

The empirical literature on music and isometric endurance is mixed, which makes the specific question of tempo -- holding genre, lyrics, and volume constant -- worth testing under a clean within-subjects design. Poon et al. (2024), using a randomized crossover trial, found that music during the bout extended time-to-exhaustion in both plank and wall-sit tasks and delayed fatigue perception without detectable differences in heart rate or blood lactate, a pattern consistent with a psychophysical rather than metabolic pathway. Feiss et al. (2021), in a between-subjects design, reported that fast- and slow-tempo music prolonged dissociative attention during a wall-sit but not during a plank, and found no tempo-driven improvement in plank endurance -- a null result likely shaped by the large between-person variance that a between-subjects design cannot absorb. The Terry et al. (2020) meta-analysis of 139 studies reported a generalized music effect of Hedges' g = 0.31 on physical performance, while Centala et al. (2020) found that fast-tempo music raised the EMG-based fatigue threshold and maximum power output without increasing end-exercise perceived exertion, consistent with improved tolerance to discomfort. Ballmann (2021) further emphasizes that preference and hedonic response moderate effect sizes, which motivated our use of genre-consistent instrumental EDM tracks on both arms. Two mechanisms are plausible for isometric tasks specifically: psychological dissociation, in which fast external stimulation competes with interoceptive fatigue cues and lowers perceived exertion, and physiological arousal, in which sympathetic activation raises motivation and tolerance for discomfort near task failure.

Our pre-registered primary hypothesis was directional: high-tempo music would increase plank time-to-volitional-exhaustion relative to low-tempo music. Two secondary hypotheses follow from the mechanisms above. Under the arousal pathway, the high-tempo condition should produce higher self-reported motivation and perceived energy. Under the dissociation pathway, the high-tempo condition should not produce proportionally higher perceived exertion despite longer holds, consistent with music competing with fatigue signals rather than simply inducing participants to try harder. These secondary outcomes are therefore tests of mechanism rather than of overall performance.

# II. Experimental Details

The treatment condition was defined as high-tempo instrumental EDM at or above 125 BPM, while the control condition was low-tempo instrumental EDM at or below 80 BPM. Participants received fixed listening links rather than choosing their own songs, were instructed to use headphones, and were asked to keep playback near a common reference volume of roughly 60 percent. The participant-facing app kept the labels neutral during the sessions, while the analysis pipeline mapped the underlying tracks into the high-tempo and low-tempo treatment definitions. This standardization kept the treatment as consistent as possible across participants in a remote setting.

**Audio stimuli.** The study used two fixed YouTube audio sources, one assigned to the high-tempo arm and one to the low-tempo arm. The high-tempo stimulus was *Best Uplifting Trance Mix - All Time Favourites #1* by The Grand Sound, streamed from https://youtu.be/iXAbte4QXKs?si=3iLV6OhMxesZBYtZ. The low-tempo stimulus was *Easy Pop Backing Track Without Bass* by Gonella, MusicArt Rapallo, streamed from https://youtu.be/x8WbHWwZixk?si=HnDa6uBvlK6j-kcc. During data collection, the app presented these tracks under neutral session labels (Audio A and Audio B) so tempo was not revealed in the participant interface; the preprocessing pipeline then mapped the stored track identifiers to the `high_tempo` and `low_tempo` analysis labels used throughout this report.

Data acquisition was conducted through a custom mobile-first web app rather than a survey-only workflow. Participants registered through the app, received direct session links, completed guided plank sessions with embedded audio and pre/post task questions, and had their responses written through a Google Apps Script backend into Google Sheets. The app was also built with bilingual English and Spanish participant flows, which made the study more accessible and helped include participants across different locales and countries rather than limiting the interface to a single language. The participant roster stored non-PII baseline metadata such as age, gender, locale, and planned session timing, while the session records captured assigned audio condition, plank duration, and survey responses. The app also offered an optional image-capture feature that, for participants who enabled it, generated a contact-sheet style image stream that was used for plank-form quality control and to help validate timing. The preprocessing pipeline then merged those records into a participant-level wide file and the canonical derived datasets used by the analysis reports.

The pilot and design logic pointed in two directions at once. First, natural baseline variance in plank endurance is large enough that a between-subjects design would be noisy and inefficient. Second, repeated isometric effort creates obvious risks of fatigue, learning, and carryover across sessions. The final experiment therefore used a randomized crossover design in which Session 2 was scheduled within a 24 to 72 hour window after Session 1, while the final crossover analysis excluded cases where the realized elapsed time between recorded sessions was under 24 hours. That choice was intended to preserve the efficiency benefits of within-person comparison while reducing temporal spillovers from the first plank attempt into the second.

## Ex Ante Power Analysis

Before data collection, we conducted a simulation-based power analysis grounded in prior literature and calibrated with pilot data. The assumed baseline plank hold time was 111 seconds with a population standard deviation of 43 seconds, drawn from Feiss et al. (2021) and consistent with the 40.9 to 58.3 second range reported in Poon et al. (2024). We simulated three literature-based effect-size scenarios spanning the range of prior findings: a conservative +5 second effect reflecting the near-null results in Feiss et al. (2021) and Centala et al. (2020), a moderate +13 second effect corresponding to the generalized Hedges' g = 0.31 reported in the Terry et al. (2020) meta-analysis of 139 studies, and an optimistic +26 second effect drawn from the Poon et al. (2024) crossover trial (Cohen's d = 0.76). We then ran a pilot-calibrated simulation using our own eight-participant pilot, which produced a within-person paired effect of +15.6 seconds and a paired-difference standard deviation of 27.65 seconds. Under the pilot-calibrated crossover simulation the within-subjects design was anticipated to reach 80 percent power at approximately 20 participants and to exceed 93 percent power at 40 participants, whereas a parallel between-subjects design would have required roughly 260 participants to reach 80 percent power under the optimistic scenario. We targeted a sample of approximately 46 participants, a size at which the crossover design was expected to deliver greater than 90 percent power if the true effect matched the pilot or Poon et al. benchmarks, while remaining informative under the Terry et al. meta-analytic scenario.

## Randomization and Balance

Treatment-order assignment was implemented through the registration app, which randomized each eligible participant to one of the two crossover sequences (high-tempo first or low-tempo first). Balance on observed baseline characteristics was not perfect. The high-tempo-first arm was on average approximately nine years younger than the low-tempo-first arm (mean age 31.2 versus 40.3 years; see Table 20 in Section VI), a meaningful imbalance in a sample of 46 participants. We treat this as a randomization deviation to flag rather than as a source of primary bias for three reasons. First, the primary crossover estimator compares each participant to themselves, so stable participant-level characteristics like age do not enter the within-person contrast and therefore cannot bias the treatment estimate on that pathway. Second, the first-period benchmark in Section VI (Model B2) includes age as a baseline covariate, which addresses the same imbalance directly in the weaker between-person design. Third, the heterogeneity analysis in Section VII (Models H1 through H4) finds no evidence that the treatment effect varies with age after Bonferroni correction, which is the nearest substantive test of whether the imbalance materially affects the conclusions. The imbalance is therefore flagged and accommodated rather than left as an unaddressed threat.

# Model Roadmap

```{r model-roadmap}
kable(model_roadmap, caption = "Model roadmap for the final report")
```

# III. Participant Flow and Attrition

The analysis begins with the full registered sample and only then narrows to the retained analytic populations. That ordering keeps attrition, missingness, and protocol exclusions visible before any treatment effects are interpreted.

```{r section-01, results='asis'}
include_section(
  "01_analysis_populations.Rmd",
  drop_sections = c("Build paired analytic dataset", "Build long repeated-measures dataset", "Write derived outputs")
)
```

Taken together, these outputs show a clear progression from `r nrow(flow)` registered participants to `r flow[, sum(completed_both_observed)]` completed-both observations and finally to `r flow[, sum(in_primary_analysis)]` protocol-valid pairs. In Table G, the `r nrow(flow) - flow[, sum(in_primary_analysis)]` non-compliers were participants who either registered but did not complete any study session or completed Session 1 but did not return for Session 2. In this study, therefore, non-compliance primarily reflects attrition within the two-session crossover protocol rather than refusal of the assigned condition itself. We did not systematically collect formal reasons for attrition, so any explanation remains descriptive rather than definitive. Based on participant communications and team observation, the most plausible reasons were failure to allocate time to complete both sessions and discomfort during or after a session that discouraged continued participation. For example, one participant reported abdominal discomfort related to a prior surgery after Session 1 and was therefore unable to complete Session 2. The attrition-bias tables further suggest that retention through both sessions was not strongly associated with randomized order or the observed baseline demographics reported there, although those checks remain descriptive rather than definitive.

# IV. Primary Crossover Estimate

The core estimate comes from the crossover sample of `r nrow(paired)` protocol-valid completed pairs. In raw paired terms, mean TTE increased by `r sprintf("%.2f", effect_mean)` seconds under high tempo, which corresponds to a `r sprintf("%.1f", percent_improvement)` percent improvement relative to the low-tempo mean. To make the specification ladder transparent, the report shows the raw paired contrast, then a participant fixed-effects model without period adjustment, and finally the period-adjusted fixed-effects crossover regression with participant-clustered standard errors.

```{r section-02, results='asis'}
include_section("02_primary_crossover_analysis.Rmd")
```

```{r stargazer-primary, results='asis'}
render_stargazer_table(
  "stargazer_primary_family",
  caption = "The `stargazer` summary below isolates the regression members of the primary family, `P2` and `P3`, so the reader can compare the fixed-effects specifications directly."
)
```

The graphical story is consistent with the regression story: most of the within-person signal leans toward longer holds under high tempo, and the center of the paired-difference distribution lies to the right of zero. The three-model ladder then shows that the estimated treatment effect remains near `r sprintf("%.2f", effect_mean)` seconds across specifications, which is exactly the pattern one would hope to see if the result is design-driven rather than specification-driven.

# V. Period and Order Diagnostics

Because the protocol required two plank attempts per participant, it is important to check whether broad session-order effects disrupted interpretation. The diagnostics below test whether performance shifted systematically from Session 1 to Session 2 and whether those patterns differed by randomized sequence arm.

```{r section-03, results='asis'}
include_section("03_period_order_diagnostics.Rmd")
```

```{r stargazer-diagnostic, results='asis'}
render_stargazer_table(
  "stargazer_diagnostic_family",
  caption = "The next `stargazer` output focuses on `D2`, the sequence diagnostic regression. It is included separately because the interaction term is the main formal check for order-specific period patterns."
)
```

These diagnostics tell two stories. First, the broad Session 2 minus Session 1 difference pooled across arms averages about `r sprintf("%.2f", unname(fe_period_ct["period_factor2", "Estimate"]))` seconds, which is small relative to the treatment effect. Second, however, the sequence diagnostic regression reveals a large and statistically significant period-by-order interaction. In a crossover design, this is the classical signature of either asymmetric carryover -- in which completing one condition alters the response under the other -- or of differential practice and fatigue between arms that is unrelated to treatment spillover. We cannot fully adjudicate between these explanations within the primary design alone. Two observations nonetheless bound the severity of the threat. First, the treatment effect points in the hypothesized direction within both randomized arms; the interaction reflects differing magnitudes of effect rather than a sign flip. Second, the first-period benchmark reported in Section VI is mechanically carryover-free by construction, because it uses only each participant's Session 1 observation and therefore cannot be contaminated by any order effect. That benchmark produces a positive point estimate in the same direction as the primary crossover result. Its standard error is much larger, as the precision comparison makes clear, but its directional agreement with the primary estimate is evidence that the main treatment signal is not purely an artifact of ordering.

# VI. First-Period Benchmark and Precision

The first-period benchmark asks what the treatment effect would have looked like if the study had been analyzed as a conventional parallel-arm trial using one independent observation per participant. That benchmark is supplementary by design and serves two purposes. Its primary purpose is to show how much precision the crossover design gained by conditioning on stable participant-level differences rather than trying to estimate the effect entirely across people. Its secondary purpose is to provide a carryover-free sensitivity check on the primary estimate: because the benchmark uses only each participant's Session 1 observation, it is mechanically immune to the period-by-order interaction flagged in Section V.

The contrast is stark in this dataset: the crossover standard error is `r sprintf("%.2f", se_pair)` seconds, whereas the first-period benchmark standard error is `r sprintf("%.2f", se_parallel)` seconds, a ratio of about `r sprintf("%.2f", precision_ratio)` to 1. The benchmark estimate therefore illustrates the design argument as much as the substantive treatment argument.

```{r section-04, results='asis'}
include_section("04_first_period_benchmark.Rmd")
```

```{r stargazer-benchmark, results='asis'}
render_stargazer_table(
  "stargazer_benchmark_family",
  caption = "The benchmark `stargazer` table below summarizes `B1` and `B2`, the two Session 1-only models. These are useful as a design contrast because they show how weak the treatment evidence becomes once the analysis stops comparing participants to themselves."
)
```

```{r section-05, results='asis'}
include_section("05_precision_power_comparison.Rmd")
```

The key takeaway is not merely that the benchmark p-value is larger; it is that the entire benchmark design is far less informative. The benchmark treatment estimate is noisy and non-significant, while the precision tables show that a parallel-arm design would have needed far more independent participants to reach the same level of precision achieved by the crossover analysis.

# VII. Exploratory Mechanisms and Heterogeneity

The secondary analyses ask why the treatment may have worked and whether it appears to work differently across subgroups. The mechanism checks focus on RPE, motivation, and energy; the heterogeneity checks examine age, gender, baseline activity level, and plank familiarity. Because subgroup exploration creates a multiple-comparisons problem, the report applies a Bonferroni correction and treats the HTE results as exploratory.

In directional terms, motivation shows the clearest secondary signal (`r sprintf("%.3f", sec_diff("Motivation"))`, p = `r sprintf("%.4f", sec_p("Motivation"))`), while the adjusted subgroup p-values remain uniformly unconvincing (`r paste(sprintf('%.3f', hte_ps_adj), collapse = ', ')`).

```{r section-06, results='asis'}
include_section("06_secondary_outcomes_and_hte.Rmd")
```

```{r stargazer-secondary, results='asis'}
render_stargazer_table(
  "stargazer_secondary_fe_family",
  caption = "This `stargazer` block summarizes the fixed-effects secondary-outcome models, `S4-S6`. It complements the paired secondary tables by showing whether the same directional patterns remain when the outcomes are re-estimated in the crossover regression framework."
)
```

```{r stargazer-hte, results='asis'}
render_stargazer_table(
  "stargazer_hte_family",
  caption = "The final exploratory `stargazer` block summarizes the heterogeneity family, `H1-H4`. Here the interaction term is the key quantity because it tests whether the treatment effect varies across the pre-specified subgroup dimension in each column."
)
```

These outputs suggest that the treatment may have worked partly through engagement or readiness to perform, because motivation moves in the expected direction more clearly than the other secondary outcomes. At the same time, the heterogeneity table does not support strong claims that the treatment was materially larger or smaller for any of the tested subgroups after Bonferroni correction.

# VIII. Design-Based Inference Check

The final inferential step asks whether the observed paired treatment effect looks unusual under the sharp null implied by the randomization mechanism itself. The bridge linear model writes the observed paired difference in regression form, while the randomization-inference procedure uses 10,000 random sign flips of the participant-level treatment differences to generate a distribution-free p-value.

```{r section-07, results='asis'}
include_section("07_randomization_inference.Rmd")
```

```{r stargazer-ri, results='asis'}
render_stargazer_table(
  "stargazer_ri_bridge_family",
  caption = "The `stargazer` output for `R1` is included here as a compact regression restatement of the paired estimand. It is not the design-based test itself, but it makes the bridge from the paired-difference statistic to an intercept-only linear model explicit."
)
```

The practical conclusion is that the observed paired effect is rare under that randomization-based null. The RI p-value of `r sprintf("%.4f", ri_two_sided)` closely reinforces the paired and fixed-effects results rather than contradicting them, which is exactly why this final check is valuable.

# IX. Discussion and Conclusion

Across the primary specifications, the treatment effect is consistent in sign and similar in magnitude. The paired estimate, the participant fixed-effects model without period adjustment, the period-adjusted fixed-effects model, and the design-based randomization-inference check all point to the same practical conclusion: in this experiment, the high-tempo condition increased plank endurance by roughly fourteen seconds relative to the low-tempo condition.

The benchmark and precision sections clarify why the design mattered so much. A first-period-only parallel-arm analysis produced a much less precise estimate, a wider confidence interval, and a non-significant result. That contrast is not a contradiction. It is evidence that the randomized crossover design was the correct tool for this question because it absorbed the large baseline variance in plank performance that would otherwise overwhelm the treatment signal.

At the same time, the study has important limits. The protocol was remote and partially self-administered, so timing quality and adherence could not be supervised in person. The treatment was also narrow: one high-tempo instrumental EDM track and one low-tempo instrumental EDM track delivered under standardized instructions. For that reason, the experiment identifies the causal effect of this bundled treatment package more cleanly than it identifies the pure effect of tempo abstracted away from song identity, production style, or genre. A different pair of tracks, or a different genre entirely such as classical or heavy metal, might produce different results.

Generalizability should therefore be stated carefully. The findings speak most directly to short-duration isometric endurance in a remote plank task, not automatically to longer cardiovascular exercise such as running or cycling. What the experiment does show is that, in this setting, a carefully standardized high-tempo music condition improved performance relative to a carefully standardized low-tempo condition, and that the crossover design was essential for estimating that effect precisely.

# References

Ballmann, C. G. (2021). The influence of music preference on exercise responses and performance: A review. *Journal of Functional Morphology and Kinesiology, 6*(2), 33. https://doi.org/10.3390/jfmk6020033

Centala, J., Pogorel, C., Pummill, S. W., & Malek, M. H. (2020). Listening to fast-tempo music delays the onset of neuromuscular fatigue. *Journal of Strength and Conditioning Research, 34*(3), 617-622. https://doi.org/10.1519/JSC.0000000000003417

Coppock, A. (n.d.). *10 things to know about statistical power*. EGAP. https://egap.org/resource/10-things-to-know-about-statistical-power/

Feiss, R., Kostrna, J., Scruggs, J. W., Pangelinan, M., & Tenenbaum, G. (2021). Effects of music tempo on perceived exertion, attention, affect, heart rate, and performance during isometric strength exercise. *Journal of Sports Sciences, 39*(2), 161-169. https://doi.org/10.1080/02640414.2020.1809974

Gerber, A. S., & Green, D. P. (2012). *Field experiments: Design, analysis, and interpretation*. W. W. Norton & Company.

Li, C., Jin, C., Zhang, Z., & Shi, P. (2022). Music recharges people: Synchronized music during aerobic exercise leads to better self-regulation performance. *PLOS ONE, 17*(12), e0278062. https://doi.org/10.1371/journal.pone.0278062

Poon, E. T.-C., Kwan, W. L., Chow, C. C., & Chan, D. K. C. (2024). Music timing differentiates fatigue perception and performance during isometric strength exercises: A crossover randomised trial. *Journal of Sports Science and Medicine, 23*, 258-264. https://doi.org/10.52082/jssm.2024.258

Terry, P. C., Karageorghis, C. I., Curran, M. L., Martin, O. V., & Parsons-Smith, R. L. (2020). Effects of music in exercise and sport: A meta-analytic review. *Psychological Bulletin, 146*(2), 91-117. https://doi.org/10.1037/bul0000216

W241 Plank Study. (2026). *DATASCI241_Music_Tempo_Effects_on_Exercise_Performance* [Computer software]. GitHub. https://github.com/karan-mids24/DATASCI241_Music_Tempo_Effects_on_Exercise_Performance.git

Angrist, J. D., & Pischke, J.-S. (2015). *Mastering 'metrics: The path from cause to effect*. Princeton University Press.
